---
title: "Assignment 9: Predicting Graduate School Admission"
author: "Jacob Janke"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center"
)

library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(forcats)
library(broom)
library(pROC)
library(car)
library(knitr)
library(scales)
```

### Introduction

The goal of this assignment is to determine which applicant characteristics are associated with graduate school admission. The outcome variable is **admit** (1 = admitted, 0 = not admitted). Predictors include:

- **gre**: GRE score
- **gpa**: Undergraduate GPA
- **rank**: Undergraduate institution rank (1 = highest tier, 4 = lowest tier)

I perform exploratory data analysis (EDA), clean and prepare the data, fit logistic regression models, evaluate model performance, and interpret results in terms of odds ratios.

### Data Loading and Cleaning

```{r load-clean-data}
# Use a relative path from Assignments/Assignment_9 to Data/
admissions_raw <- read_csv("../../Data/GradSchool_Admissions.csv", show_col_types = FALSE)

# Initial structure
str(admissions_raw)

# Cleaned analysis dataset
admissions <- admissions_raw %>%
  mutate(
    admit = as.integer(admit),
    admit_factor = factor(admit, levels = c(0, 1), labels = c("Not Admitted", "Admitted")),
    rank = factor(rank, levels = c(1, 2, 3, 4), ordered = TRUE)
  )

# Data quality checks
missing_summary <- colSums(is.na(admissions))
duplicate_rows <- sum(duplicated(admissions_raw))

missing_summary
duplicate_rows

# Basic range checks
admissions %>%
  summarise(
    min_gre = min(gre),
    max_gre = max(gre),
    min_gpa = min(gpa),
    max_gpa = max(gpa)
  )
```

**Cleaning notes:**

- No missing values were found.
- No duplicate rows were found.
- GRE and GPA ranges are reasonable.
- I converted `rank` to an ordered factor and created `admit_factor` for clear plotting labels.

### Exploratory Data Analysis

#### Summary Statistics

```{r summary-statistics}
summary(admissions)

numeric_summary <- admissions %>%
  select(admit, gre, gpa) %>%
  summarise(
    across(
      everything(),
      list(
        mean = ~mean(.),
        sd = ~sd(.),
        min = ~min(.),
        q1 = ~quantile(., 0.25),
        median = ~median(.),
        q3 = ~quantile(., 0.75),
        max = ~max(.)
      ),
      .names = "{.col}_{.fn}"
    )
  )

kable(numeric_summary, digits = 3, caption = "Numeric summary statistics for admission outcome and predictors.")

rank_counts <- admissions %>% count(rank, name = "n")
kable(rank_counts, caption = "Counts by undergraduate institution rank.")
```

#### Predictor Distributions

```{r dist-gre, fig.cap="Distribution of GRE scores."}
ggplot(admissions, aes(x = gre)) +
  geom_histogram(binwidth = 30, fill = "steelblue", color = "white", alpha = 0.9) +
  geom_density(aes(y = after_stat(count * 30)), color = "black", linewidth = 0.8) +
  labs(title = "GRE Score Distribution", x = "GRE", y = "Count") +
  theme_minimal()
```

```{r dist-gpa, fig.cap="Distribution of GPA values."}
ggplot(admissions, aes(x = gpa)) +
  geom_histogram(binwidth = 0.15, fill = "darkseagreen", color = "white", alpha = 0.9) +
  geom_density(aes(y = after_stat(count * 0.15)), color = "black", linewidth = 0.8) +
  labs(title = "GPA Distribution", x = "GPA", y = "Count") +
  theme_minimal()
```

```{r dist-rank, fig.cap="Distribution of undergraduate institution rank (1 = top tier)."}
ggplot(admissions, aes(x = rank)) +
  geom_bar(fill = "mediumpurple", alpha = 0.9) +
  labs(title = "Institution Rank Distribution", x = "Institution Rank", y = "Count") +
  theme_minimal()
```

#### Relationships Between Predictors and Admission

```{r relation-gre-admit, fig.cap="GRE by admission status with boxplot and jittered points."}
ggplot(admissions, aes(x = admit_factor, y = gre, fill = admit_factor)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.35) +
  scale_fill_manual(values = c("tomato", "steelblue")) +
  labs(title = "GRE by Admission Outcome", x = "Admission Outcome", y = "GRE") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r relation-gpa-admit, fig.cap="GPA by admission status with boxplot and jittered points."}
ggplot(admissions, aes(x = admit_factor, y = gpa, fill = admit_factor)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.35) +
  scale_fill_manual(values = c("tomato", "steelblue")) +
  labs(title = "GPA by Admission Outcome", x = "Admission Outcome", y = "GPA") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r relation-rank-admit, fig.cap="Admission rates by undergraduate institution rank."}
admit_by_rank <- admissions %>%
  group_by(rank) %>%
  summarise(admit_rate = mean(admit), n = n(), .groups = "drop")

kable(admit_by_rank, digits = 3, caption = "Observed admission rate by institution rank.")

ggplot(admissions, aes(x = rank, fill = admit_factor)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("tomato", "steelblue")) +
  labs(
    title = "Proportion Admitted by Institution Rank",
    x = "Institution Rank",
    y = "Proportion",
    fill = "Outcome"
  ) +
  theme_minimal()
```

#### Correlation Analysis

```{r correlation-analysis, fig.cap="Correlation heatmap among numeric variables (rank converted to numeric for this analysis)."}
corr_df <- admissions %>%
  mutate(rank_numeric = as.numeric(rank)) %>%
  select(admit, gre, gpa, rank_numeric)

corr_matrix <- cor(corr_df)
kable(round(corr_matrix, 3), caption = "Correlation matrix (Pearson).")

corr_long <- as.data.frame(as.table(corr_matrix))
colnames(corr_long) <- c("Var1", "Var2", "Correlation")

ggplot(corr_long, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", Correlation)), size = 4) +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "navy", midpoint = 0) +
  labs(title = "Correlation Heatmap", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Logistic Regression Modeling

#### Fit Individual Predictor Models and Full Model

```{r fit-models}
# Logistic models
model_gre <- glm(admit ~ gre, data = admissions, family = binomial)
model_gpa <- glm(admit ~ gpa, data = admissions, family = binomial)
model_rank <- glm(admit ~ rank, data = admissions, family = binomial)
model_full <- glm(admit ~ gre + gpa + rank, data = admissions, family = binomial)
model_null <- glm(admit ~ 1, data = admissions, family = binomial)

summary(model_gre)
summary(model_gpa)
summary(model_rank)
summary(model_full)
```

#### Compare Model Fit

```{r model-comparison}
mcfadden_r2 <- function(model, null_model) {
  1 - as.numeric(logLik(model) / logLik(null_model))
}

model_comparison <- tibble(
  Model = c("GRE only", "GPA only", "Rank only", "Full model"),
  AIC = c(AIC(model_gre), AIC(model_gpa), AIC(model_rank), AIC(model_full)),
  McFadden_R2 = c(
    mcfadden_r2(model_gre, model_null),
    mcfadden_r2(model_gpa, model_null),
    mcfadden_r2(model_rank, model_null),
    mcfadden_r2(model_full, model_null)
  )
)

kable(model_comparison, digits = 3, caption = "Model comparison using AIC and McFadden pseudo-R².")
```

#### Coefficient Interpretation and Odds Ratios (Full Model)

```{r odds-ratios-full-model}
odds_table <- tidy(model_full, conf.int = TRUE, exponentiate = TRUE)

kable(
  odds_table,
  digits = 3,
  caption = "Full model odds ratios with 95% confidence intervals."
)
```

Interpretation guide:

- An odds ratio (OR) > 1 indicates increased odds of admission as the predictor increases.
- An OR < 1 indicates decreased odds of admission.
- For factor levels of `rank`, each level is interpreted relative to rank 1 (top-tier institutions).

### Model Performance and Diagnostics

#### Classification Metrics and ROC Curve

```{r performance-metrics, fig.cap="ROC curve for the full logistic regression model."}
# Predicted probabilities and classes
admissions <- admissions %>%
  mutate(
    pred_prob_full = predict(model_full, type = "response"),
    pred_class_full = if_else(pred_prob_full >= 0.5, 1L, 0L)
  )

# Confusion matrix
conf_mat <- table(Predicted = admissions$pred_class_full, Actual = admissions$admit)
conf_mat

# Manual performance metrics
TP <- conf_mat["1", "1"]
TN <- conf_mat["0", "0"]
FP <- conf_mat["1", "0"]
FN <- conf_mat["0", "1"]

accuracy <- (TP + TN) / sum(conf_mat)
sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)
precision <- TP / (TP + FP)

performance_tbl <- tibble(
  Metric = c("Accuracy", "Sensitivity (Recall)", "Specificity", "Precision"),
  Value = c(accuracy, sensitivity, specificity, precision)
)

kable(performance_tbl, digits = 3, caption = "Classification metrics at a 0.50 threshold.")

# ROC + AUC
roc_obj <- roc(admissions$admit, admissions$pred_prob_full, quiet = TRUE)
auc_value <- as.numeric(auc(roc_obj))
auc_value

plot(roc_obj, col = "blue", lwd = 2, main = paste("ROC Curve (AUC =", round(auc_value, 3), ")"))
abline(a = 0, b = 1, lty = 2, col = "gray40")
```

#### Additional Diagnostics

```{r diagnostics}
# 1) Multicollinearity diagnostics
vif_values <- vif(model_full)
vif_values

# 2) Deviance residual summary (fit check)
resid_summary <- summary(residuals(model_full, type = "deviance"))
resid_summary

# 3) Influential points via Cook's distance
cooks_d <- cooks.distance(model_full)

plot(
  cooks_d,
  type = "h",
  main = "Cook's Distance for Full Model",
  ylab = "Cook's distance",
  xlab = "Observation"
)
abline(h = 4 / nrow(admissions), col = "red", lty = 2)

# 4) Leverage diagnostics
plot(
  hatvalues(model_full),
  ylab = "Leverage",
  xlab = "Observation",
  main = "Leverage Values"
)
abline(h = 2 * mean(hatvalues(model_full)), col = "red", lty = 2)
```

### Conclusions

Based on the EDA and logistic regression models:

1. **GPA and GRE** are positively associated with admission probability.
2. **Institution rank matters**: applicants from lower-tier institutions (higher numeric rank) generally have lower odds of admission relative to top-tier institutions.
3. The **full model** performs better than single-predictor models by combining academic performance and institutional context.
4. Diagnostics indicate the model is reasonably stable and suitable for inference in this assignment context.

Overall, admission appears to be most strongly predicted by a combination of applicant achievement (GPA, GRE) and undergraduate institution rank.
