---
title: "Exam 3 - Skills Test"
author: "Jacob Janke"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Task 1: Load and Clean FacultySalaries_1995.csv

First, I'll load the necessary libraries and import the data.

```{r load_libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(broom)
```

```{r load_faculty_data}
# Load the faculty salaries data
faculty <- read.csv("FacultySalaries_1995.csv")

# View the structure
head(faculty)
```

Now I need to tidy this data. The issue is that we have separate columns for different ranks (Full, Assoc, Assistant) but we need them in a long format with Rank as a variable.

```{r tidy_faculty_data}
# Tidy the data by pivoting the salary columns
faculty_tidy <- faculty %>%
  select(State, Tier, AvgFullProfSalary, AvgAssocProfSalary, AvgAssistProfSalary) %>%
  pivot_longer(
    cols = starts_with("Avg"),
    names_to = "Rank",
    values_to = "Salary"
  ) %>%
  mutate(
    Rank = case_when(
      Rank == "AvgFullProfSalary" ~ "Full",
      Rank == "AvgAssocProfSalary" ~ "Associate",
      Rank == "AvgAssistProfSalary" ~ "Assistant"
    ),
    Rank = factor(Rank, levels = c("Assistant", "Associate", "Full"))
  )

head(faculty_tidy)
```

Now I'll recreate the graph from Fig1.png:

```{r fig1_recreation, fig.align='center', fig.width=10, fig.height=5}
ggplot(faculty_tidy, aes(x = Rank, y = Salary, fill = Rank)) +
  geom_boxplot() +
  facet_grid(~ Tier) +
  labs(
    x = "Rank",
    y = "Salary"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Task 2: Build ANOVA Model

Building an ANOVA model to test the influence of State, Tier, and Rank on Salary without interactions:

```{r anova_model}
# Build the ANOVA model
anova_model <- aov(Salary ~ State + Tier + Rank, data = faculty_tidy)

# Display the summary
summary(anova_model)
```

The ANOVA results show that State, Tier, and Rank all have significant effects on faculty salaries.

---

# Task 3: Load and Tidy Juniper_Oils.csv

Now loading the Juniper oils dataset:

```{r load_juniper_data}
# Load the data
juniper <- read.csv("Juniper_Oils.csv")

# View the structure
head(juniper)
```

The data needs to be tidied - we need to convert the chemical compound columns into a long format:

```{r tidy_juniper_data}
# List of chemical compounds (as they appear in the data with dots)
chemicals_in_data <- c("alpha.pinene","para.cymene","alpha.terpineol","cedr.9.ene",
                       "alpha.cedrene","beta.cedrene","cis.thujopsene","alpha.himachalene",
                       "beta.chamigrene","cuparene","compound.1","alpha.chamigrene",
                       "widdrol","cedrol","beta.acorenol","alpha.acorenol",
                       "gamma.eudesmol","beta.eudesmol","alpha.eudesmol","cedr.8.en.13.ol",
                       "cedr.8.en.15.ol","compound.2","thujopsenal")

# Original chemical names with hyphens for display
chemicals_display <- c("alpha-pinene","para-cymene","alpha-terpineol","cedr-9-ene",
                       "alpha-cedrene","beta-cedrene","cis-thujopsene","alpha-himachalene",
                       "beta-chamigrene","cuparene","compound 1","alpha-chamigrene",
                       "widdrol","cedrol","beta-acorenol","alpha-acorenol",
                       "gamma-eudesmol","beta-eudesmol","alpha-eudesmol","cedr-8-en-13-ol",
                       "cedr-8-en-15-ol","compound 2","thujopsenal")

# Create a named vector for mapping
chem_mapping <- setNames(chemicals_display, chemicals_in_data)

# Tidy the data
juniper_tidy <- juniper %>%
  select(YearsSinceBurn, all_of(chemicals_in_data)) %>%
  pivot_longer(
    cols = all_of(chemicals_in_data),
    names_to = "ChemicalID",
    values_to = "Concentration"
  ) %>%
  mutate(ChemicalID = chem_mapping[ChemicalID])

head(juniper_tidy)
```

---

# Task 4: Create Graph of YearsSinceBurn vs Concentration

Creating a faceted graph showing the relationship between years since burn and chemical concentrations:

```{r fig2_recreation, fig.align='center', fig.width=12, fig.height=14}
ggplot(juniper_tidy, aes(x = YearsSinceBurn, y = Concentration)) +
  geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "grey80") +
  facet_wrap(~ ChemicalID, scales = "free_y", ncol = 5) +
  labs(
    x = "YearsSinceBurn",
    y = "Concentration"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 9),
    axis.text = element_text(size = 7),
    panel.grid = element_line(color = "grey90")
  )
```

---

# Task 5: GLM Model to Find Significant Chemicals

Building a generalized linear model to test which chemicals are significantly affected by years since burn:

```{r glm_model}
# Build a GLM model with interaction between ChemicalID and YearsSinceBurn
glm_model <- lm(Concentration ~ ChemicalID * YearsSinceBurn, data = juniper_tidy)

# Use broom to tidy the model output
glm_tidy <- tidy(glm_model)

# Find chemicals with significant interaction with YearsSinceBurn
significant_interactions <- glm_tidy %>%
  filter(grepl(":YearsSinceBurn", term) & p.value < 0.05)

# Extract the chemical names from significant interactions
sig_chem_names <- gsub("ChemicalID", "", 
                       gsub(":YearsSinceBurn", "", significant_interactions$term))

# Get all terms (main effects and interactions) for these significant chemicals
significant_chemicals <- glm_tidy %>%
  filter(
    term %in% paste0("ChemicalID", sig_chem_names) |
    term %in% paste0("ChemicalID", sig_chem_names, ":YearsSinceBurn")
  ) %>%
  mutate(term = gsub("ChemicalID", "", term)) %>%
  arrange(p.value) %>%
  select(term, estimate, std.error, statistic, p.value)

# Display the significant chemicals
print(significant_chemicals, n = Inf)
```

The table above shows the chemicals that have concentrations significantly affected by the number of years since the burn event. These chemicals show either significant increases or decreases in concentration over time, as indicated by their coefficient estimates and p-values less than 0.05.

---

# Summary

This exam covered:

1. **Data tidying**: Transformed the FacultySalaries dataset from wide to long format to enable proper analysis and visualization
2. **ANOVA modeling**: Built a model to test the effects of State, Tier, and Rank on faculty salaries
3. **Data wrangling**: Tidied the Juniper oils dataset by converting multiple chemical columns into a long format
4. **Data visualization**: Created faceted plots to visualize trends in chemical concentrations over time
5. **Statistical modeling**: Used GLM to identify which chemicals show significant changes in concentration based on years since burn

All analyses were completed using relative file paths within the R project structure.
